---
title: "Advent of Code: Days 21-25"
description: |
  My solutions to the #AdventOfCode2021 coding challenges, days 21 through 25.
author:
  - name: Taylor Dunn
date: 2021-12-21
params:
  date: 2021-12-21
  slug: "advent-of-code-2021-days-21-25"
categories:
  - Advent of Code
  - Python
  - R
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r setup, include=TRUE, code_folding="R setup"}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(gt)
library(tictoc)
library(lubridate)

library(dunnr)
extrafont::loadfonts(device = "win", quiet = TRUE)
theme_set(theme_td_grid())
set_geom_fonts()
set_palette()
```

<details><summary>Python setup</summary>

```{r py-setup1}
Sys.setenv(
  RETICULATE_PYTHON = here("..", "python", ".venv", "Scripts", "python.exe")
)
library(reticulate)
```

```{python py-setup2}
import numpy as np
import pandas as pd
import timeit
```

</details>

## Day 21: Dirac Dice

```{r}
day21 <- read_lines(here("_posts", "2021-12-21-advent-of-code-2021-days-21-25",
                         "day21-input.txt"))
day21 %>% str_trunc(80)
```

### Part 1

>
There's not much to do as you slowly descend to the bottom of the ocean. The submarine computer challenges you to a nice game of Dirac Dice.
>
This game consists of a single die, two pawns, and a game board with a circular track containing ten spaces marked `1` through `10` clockwise. Each player's starting space is chosen randomly (your puzzle input). Player 1 goes first.
>
Players take turns moving. On each player's turn, the player rolls the die three times and adds up the results. Then, the player moves their pawn that many times forward around the track (that is, moving clockwise on spaces in order of increasing value, wrapping back around to `1` after `10`). So, if a player is on space `7` and they roll `2`, `2`, and `1`, they would move forward 5 times, to spaces `8`, `9`, `10`, `1`, and finally stopping on `2`.
>
After each player moves, they increase their score by the value of the space their pawn stopped on. Players' scores start at `0`. So, if the first player starts on space `7` and rolls a total of `5`, they would stop on space `2` and add `2` to their score (for a total score of `2`). The game immediately ends as a win for any player whose score reaches at least `1000`.
>
Since the first game is a practice game, the submarine opens a compartment labeled deterministic dice and a 100-sided die falls out. This die always rolls `1` first, then `2`, then `3`, and so on up to `100`, after which it starts over at 1 again. Play using this die.
>
The moment either player wins, what do you get if you multiply the score of the losing player by the number of times the die was rolled during the game?

```{r}
player1 <- str_extract(day21[1], "position: \\d") %>% readr::parse_number()
player2 <- str_extract(day21[2], "position: \\d") %>% readr::parse_number()
player1; player2
```

Define a custom modulo function that returns 1-10 instead of 0-9:

```{r}
custom_mod <- function(val) {
  ((val - 1) %% 10) + 1
}
```

Loop until a winner is found:

```{r}
player1_score <- player2_score <- dice <- turn <- 0

while (player1_score < 1000 & player2_score < 1000) {
  turn <- turn + 1
  dice <- dice + 3
  
  roll <- sum(seq(dice, dice - 2))
  if (turn %% 2 == 1) {
    player1 <- custom_mod(player1 + roll)
    player1_score <- player1_score + player1
  } else {
    player2 <- custom_mod(player2 + roll)
    player2_score <- player2_score + player2
  }
}
player1_score; player2_score
```

Multiply the turn number by 3 to get the number of dice rolls, then multiply that by the losing players score:

```{r}
(turn * 3) * player2_score
```

Python:

```{python}
player1 = int(r.day21[0][-1])
player2 = int(r.day21[1][-1])

player1_score, player2_score, dice, turn = 0, 0, 0, 0

while player1_score < 1000 and player2_score < 1000:
  turn += 1
  dice += 3
  
  roll = sum(range(dice - 2, dice + 1))
  if turn % 2 == 1:
    player1 = ((player1 + roll - 1) % 10) + 1
    player1_score += player1
  else:
    player2 = ((player2 + roll - 1) % 10) + 1
    player2_score += player2

turn * 3 * player2_score
```

### Part 2

>
Now that you're warmed up, it's time to play the real game.
>
A second compartment opens, this time labeled Dirac dice. Out of it falls a single three-sided die.
>
As you experiment with the die, you feel a little strange. An informational brochure in the compartment explains that this is a quantum die: when you roll it, the universe splits into multiple copies, one copy for each possible outcome of the die. In this case, rolling the die always splits the universe into three copies: one where the outcome of the roll was `1`, one where it was `2`, and one where it was `3`.
>
The game is played the same as before, although to prevent things from getting too far out of hand, the game now ends when either player's score reaches at least `21`.

On each turn, the Dirac dice is rolled three times, leading to $3 \times 3 \times 3 = 27$ permutations, i.e. 27 universes.
However, the sum of these three rolls can only take on the following values:

```{r}
roll_perm <- crossing(r1 = 1:3, r2 = 1:3, r3 = 1:3) %>%
  mutate(roll = r1 + r2 + r3) %>%
  count(roll)
roll_perm
```

Define a function to roll the dice for a single player, but keep track of the *states* of the game (instead of each game individually), where a state is uniquely identified by the positions and scores of players 1 and 2.

```{r}
roll_dirac <- function(players, player = 1) {
  players <- players %>% crossing(roll_perm)
  if (player == 1) {
    players <- players %>%
      mutate(pos1 = custom_mod(pos1 + roll), score1 = score1 + pos1,
             n_univ = n_univ * n)
  } else {
    players <- players %>%
      mutate(pos2 = custom_mod(pos2 + roll), score2 = score2 + pos2,
             n_univ = n_univ * n)
  }
  players %>%
      group_by(pos1, score1, pos2, score2) %>%
      summarise(n_univ = sum(n_univ), .groups = "drop")
}

# Sample input
players <- tibble(pos1 = 4, score1 = 0, pos2 = 8, score2 = 0, n_univ = 1)
# After two turns
players %>% roll_dirac(player = 1) %>% roll_dirac(player = 2)
```

After each roll, the number of universes (`n_univ`) is updated for each state.

Then a function to find the winners in each universe.
The loop works by removing winning states (`winner_universes`) from the active states (`players`) until none remain:

```{r}
find_winners <- function(players) {
  turn <- 0
  # Empty tibble to compile winners
  winner_universes <- players %>% filter(FALSE)
  
  while (nrow(players) > 0) {
    turn <- turn + 1
    
    players <- roll_dirac(players,
                          player = ifelse(turn %% 2 == 1, 1, 2))
    
    winners <- (players$score1 > 20) | (players$score2 > 20)
    if (any(winners)) {
      winner_universes <- winner_universes %>%
        bind_rows(players %>% filter(winners))
      
      players <- players %>% filter(!winners)
    }
  }
  winner_universes
}

winner_universes <- find_winners(players)

winner_universes %>%
  mutate(winner = ifelse(score1 > 20, "Player 1", "Player 2")) %>%
  group_by(winner) %>%
  summarise(n_univ = sum(n_univ), .groups = "drop") %>%
  mutate(n_univ = format(n_univ, scientific = FALSE))
```

Now apply it to my input to find the puzzle answer:

```{r}
players <- tibble(pos1 = player1, score1 = 0, pos2 = player2, score2 = 0,
                  n_univ = 1)
winner_universes <- find_winners(players)

winner_universes %>%
  mutate(winner = ifelse(score1 > 20, "Player 1", "Player 2")) %>%
  group_by(winner) %>%
  summarise(n_univ = sum(n_univ), .groups = "drop") %>%
  mutate(n_univ = format(n_univ, scientific = FALSE))
```

Python with `numpy` and `pandas`:

```{python}
from itertools import product
roll_perm = list(product([1, 2, 3], [1, 2, 3], [1, 2, 3]))
roll_perm = [sum(rolls) for rolls in roll_perm]
roll_perm = pd.DataFrame(roll_perm, columns = ['roll']) \
  .value_counts().to_frame('n').reset_index()
  
def roll_dirac(players, player = 1):
  if player == 1:
    players = players.merge(roll_perm, how = 'outer', on = 'join_key')
    players = players \
      .assign(pos1 = (players['pos1'] + players['roll'] - 1) % 10 + 1)
    players = players \
      .assign(score1 = players['score1'] + players['pos1'],
              n_univ = players['n_univ'] * players['n'])
  else:
    players = players.merge(roll_perm, how = 'outer', on = 'join_key')
    players = players \
      .assign(pos2 = (players['pos2'] + players['roll'] - 1) % 10 + 1)
    players = players \
      .assign(score2 = players['score2'] + players['pos2'],
              n_univ = players['n_univ'] * players['n'])
    
  players = players \
    .groupby(['pos1', 'score1', 'pos2', 'score2', 'join_key']) \
    .agg({'n_univ': 'sum'}).reset_index()
  return players

def find_winners(players):
  turn = 0
  winner_universes = players.drop(players.index)
  
  while len(players) > 0:
    turn += 1
    
    player = 1 if (turn % 2 == 1) else 2
    players = roll_dirac(players, player)
    
    winners = (players['score1'].values > 20) | (players['score2'].values > 20)
    if any(winners):
      winner_universes = winner_universes.append(players.loc[winners])
      players = players.loc[~winners]
  
  return winner_universes

players = pd.DataFrame({'pos1': player1, 'pos2': player2,
                        'score1': 0, 'score2': 0, 'n_univ': 1}, index = [0])
  
# I'm not aware of a pandas function similar to tidyr::crossing, so use a hacky
#  join_key column
roll_perm['join_key'] = 0
players['join_key'] = 0

winner_universes = find_winners(players)

winner_universes \
  .assign(winner = np.where(winner_universes['score1'] > 20,
                            'Player 1', 'Player 2')) \
  .groupby(['winner']) \
  .agg({'n_univ': 'sum'})
```

## Day 22: Reactor Reboot

```{r}
day22 <- read_lines(here("_posts", "2021-12-21-advent-of-code-2021-days-21-25",
                         "day22-input.txt"))
day22 %>% head()
```

### Part 1

>
Operating at these extreme ocean depths has overloaded the submarine's reactor; it needs to be rebooted.
>
The reactor core is made up of a large 3-dimensional grid made up entirely of cubes, one cube per integer 3-dimensional coordinate (`x,y,z`). Each cube can be either on or off; at the start of the reboot process, they are all off. (Could it be an old model of a reactor you've seen before?)
>
To reboot the reactor, you just need to set all of the cubes to either on or off by following a list of reboot steps (your puzzle input). Each step specifies a cuboid (the set of all cubes that have coordinates which fall within ranges for `x`, `y`, and `z`) and whether to turn all of the cubes in that cuboid on or off.
>
The initialization procedure only uses cubes that have `x`, `y`, and `z` positions of at least `-50` and at most `50`. For now, ignore cubes outside this region.
>
Execute the reboot steps. Afterward, considering only cubes in the region `x=-50..50,y=-50..50,z=-50..50`, how many cubes are on?

Parse the reboot steps:

```{r}
reboot_steps <- tibble(x = day22) %>%
  extract(
    col = x, into = c("on", "xmin", "xmax", "ymin", "ymax", "zmin", "zmax"),
    regex = "(.*) x=(.*)\\.\\.(.*),y=(.*)\\.\\.(.*),z=(.*)\\.\\.(.*)",
    convert = TRUE
  ) %>%
  # Use a logical to indicate on/off
  mutate(on = on == "on", step_num = row_number())
reboot_steps
```

For part 1, I only need the steps within -50 and +50 in each dimension:

```{r}
reboot_steps_part1 <- reboot_steps %>%
  filter(xmax >= -50, ymax >= -50, zmax >= -50,
         xmin <= 50, ymin <= 50, zmin <= 50)
```

Represent the 100x100x100 grid of cubes with an `array` and a function to update it from a single step:

```{r}
cubes <- array(FALSE, dim = c(100, 100, 100))

# For my own convenience, add 50 to each range so that I can index the array
#  with 1-100
reboot_steps_part1 <- reboot_steps_part1 %>%
  mutate(across(xmin:zmax, ~ . + 50))

switch_cubes <- function(cubes, reboot_step) {
  cubes[reboot_step$xmin:reboot_step$xmax,
        reboot_step$ymin:reboot_step$ymax,
        reboot_step$zmin:reboot_step$zmax] <- reboot_step$on
  return(cubes)
}
```

Now apply it and take the sum to get the number of `on` cubes:

```{r}
for (reboot_step in reboot_steps_part1 %>% split(.$step_num)) {
  cubes <- switch_cubes(cubes, reboot_step)
}
sum(cubes)
```

### Part 2

>
Now that the initialization procedure is complete, you can reboot the reactor.
>
Starting again with all cubes off, execute all reboot steps. Afterward, considering all cubes, how many cubes are on?

My solution to part 1 won't do here.

```{r}
reboot_steps %>%
  summarise(across(c(xmin, ymin, zmin), min),
            across(c(xmax, ymax, zmax), max))
```

Can't really work with a 20k x 20k x 20k grid in R.

Instead, for the cuboid defined in each step, I will consider every cuboid that came before it.
If I find any intersecting cuboids, I will separate them into smaller cuboids that don't intersect.
For the on/off instruction of each cuboid, I will keep the last instruction.

```{r}
get_cuboids <- function(reboot_steps) {
  cuboids <- tibble(xmin = numeric(), xmax = numeric(),
                    ymin = numeric(), ymax = numeric(),
                    zmin = numeric(), zmax = numeric())
  
  n_steps <- nrow(reboot_steps)
  for (cuboid1 in reboot_steps %>% split(.$step_num)) {
    message(paste0("Step ", cuboid1$step_num, " of ", n_steps))
    
    # After a lot of trial and error with the example input, I found out that I
    #  had an off-by-one error that was fixed by nudging the upper bounds
    cuboid1 <- cuboid1 %>%
      mutate(xmax = xmax + 1, ymax = ymax + 1, zmax = zmax + 1)
    
    # Declare an empty tibble to compile new cuboids
    new_cuboids <- cuboids %>% filter(FALSE)
    # Loop over the cuboids gathered so far
    for (i in seq_len(nrow(cuboids))) {
      cuboid2 <- cuboids %>% slice(i)
      
      # Check if the new cuboid overlaps with the old cuboid
      x_overlap <- (cuboid1$xmax > cuboid2$xmin) & (cuboid1$xmin < cuboid2$xmax)
      y_overlap <- (cuboid1$ymax > cuboid2$ymin) & (cuboid1$ymin < cuboid2$ymax)
      z_overlap <- (cuboid1$zmax > cuboid2$zmin) & (cuboid1$zmin < cuboid2$zmax)
      
      if (x_overlap & y_overlap & z_overlap) {
        # If the left edge of cuboid2 is to the left of cuboid1
        if (cuboid2$xmin < cuboid1$xmin) {
          # Slice off that portion into a new cuboid
          new_cuboids <- new_cuboids %>%
            bind_rows(cuboid2 %>% mutate(xmax = cuboid1$xmin))
          # And shrink cuboid2, as the new cuboid covers that area already
          cuboid2 <- cuboid2 %>% mutate(xmin = cuboid1$xmin)
        }
        # Do the same for the other 5 directions
        if (cuboid2$xmax > cuboid1$xmax) {
          new_cuboids <- new_cuboids %>%
            bind_rows(cuboid2 %>% mutate(xmin = cuboid1$xmax))
          cuboid2 <- cuboid2 %>% mutate(xmax = cuboid1$xmax)
        }
        if (cuboid2$ymin < cuboid1$ymin) {
          new_cuboids <- new_cuboids %>%
            bind_rows(cuboid2 %>% mutate(ymax = cuboid1$ymin))
          cuboid2 <- cuboid2 %>% mutate(ymin = cuboid1$ymin)
        }
        if (cuboid2$ymax > cuboid1$ymax) {
          new_cuboids <- new_cuboids %>%
            bind_rows(cuboid2 %>% mutate(ymin = cuboid1$ymax))
          cuboid2 <- cuboid2 %>% mutate(ymax = cuboid1$ymax)
        }
        if (cuboid2$zmin < cuboid1$zmin) {
          new_cuboids <- new_cuboids %>%
            bind_rows(cuboid2 %>% mutate(zmax = cuboid1$zmin))
          cuboid2 <- cuboid2 %>% mutate(zmin = cuboid1$zmin)
        }
        if (cuboid2$zmax > cuboid1$zmax) {
          new_cuboids <- new_cuboids %>%
            bind_rows(cuboid2 %>% mutate(zmin = cuboid1$zmax))
          cuboid2 <- cuboid2 %>% mutate(zmax = cuboid1$zmax)
        }
      }
      else {
        new_cuboids <- new_cuboids %>% bind_rows(cuboid2)
      }
    }
    new_cuboids <- new_cuboids %>% bind_rows(cuboid1)
    cuboids <- new_cuboids
  }
  return(cuboids)
}
```

As a demonstration, consider the first 3 steps (which happen to have overlapping cuboids) in just 2 dimensions:

```{r fig.height=3, fig.width=6}
reboot_steps3 <- reboot_steps %>% slice(1:3)

p <- reboot_steps3 %>%
  mutate(step_num = factor(step_num)) %>%
  ggplot(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax)) +
  geom_rect(aes(fill = step_num),alpha = 0.5)
p
```

These are the resulting cuboids after splitting them at the intersections:

```{r fig.height=3, fig.width=6, message=FALSE, preview=TRUE}
p +
  geom_rect(
    data = get_cuboids(reboot_steps3) %>%
      # Shrink the areas slightly to visualize the borders
      mutate(across(c(xmin, ymin), ~ .x + 0.5),
             across(c(xmax, ymax), ~ .x - 0.5)),
    fill = NA, color = "black")
```

Now run it on the full input (and time it because I expect it to take a long time):

```{r part2_loop, cache=TRUE, message=FALSE}
tic()
cubes <- get_cuboids(reboot_steps)
toc()
```

~15 minute run time.

Now I have a list of cuboids which are non-overlapping.
I can filter down to just the ones that are `on` and count their volumes to get the total number of on cubes:

```{r}
cubes %>%
  filter(on) %>%
  mutate(volume = (xmax - xmin) * (ymax - ymin) * (zmax - zmin)) %>%
  summarise(n_on = sum(volume) %>% format(scientific = FALSE))
```

## Reproducibility {.appendix}

<details><summary>Session info</summary>

```{r echo=FALSE}
devtools::session_info()$platform
devtools::session_info()$packages %>%
  rmarkdown::paged_table()
```

</details>

<details><summary>Git repository</summary>

```{r echo=FALSE}
git2r::repository()
```

</details>

```{r echo=FALSE}
dunnr::get_distill_source(date = params$date, slug = params$slug)
```

