---
title: "Ordinal regression in R: part 2"
description: |
  A theoretical and applied walkthrough of ordinal regression.
  Part 2: Bayesian approach with `brms`.
author:
  - name: Taylor Dunn
date: 2020-03-17
params:
  date: 2020-03-17
  slug: 
categories:
  - regression
  - ordinal
  - Bayesian statistics
output:
  distill::distill_article:
    self_contained: false
    toc: true
bibliography: references.bib
---

```{r setup, include=TRUE, code_folding="Setup"}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dunnr)
library(gt)
library(broom)
library(broom.mixed)
library(patchwork)

extrafont::loadfonts(device = "win", quiet = TRUE)
theme_set(theme_td())
set_geom_fonts()
set_palette()

wine_red <- "#58181F"
update_geom_defaults("point", list(color = wine_red))
update_geom_defaults("line", list(color = wine_red))
```

This is part 2 of me learning ordinal regression in R.
Previously, I took the [frequentist approach with the `ordinal` package.](2020-03-15-ordinal-regression-in-r-part-1/)
Here, I'll use [`brms` package](https://cran.r-project.org/web/packages/brms/index.html) to fit Bayesian mixed models via Stan.

Though I won't be reproducing their examples, @Burkner2019 give a great tutorial of using `brms` for ordinal regression models.
It also frames the cumulative model in the terms of a latent (not observable) continuous variable $\tilde{y}$, which has been categorized into the observed ordinal variable $y$.
I found this way of thinking very intuitive, and it helped make a lot of the concepts click.

This post also serves as practice in Bayesian inference, so I'll be comparing the results here to those from part 1, and explore different choices of prior distributions.

## Setup

Load `brms`, `tidybayes` and the `wine` data from @Randall1989 that was analyzed in part 1.

```{r}
library(brms)
library(tidybayes)
# Detect and set the number of cores for MCMC
options(mc.cores = parallel::detectCores())

library(ordinal)
data(wine)
wine <- as_tibble(wine)
```

## Fitting the models

I will be fitting these models of `wine` bitterness `rating`s:

$$
\begin{align}
\text{logit}(p(y_i \leq j)) &= \theta_j - u( \text{judge}_i) \\
\text{logit}(p(y_i \leq j)) &= \theta_j - \beta_1 \text{temp}_i - \beta_2 \text{contact}_i - u( \text{judge}_i) \\
\end{align}
$$

where $p(y_i \leq j)$ is the probability of a `rating` less than or equal to $j$, $\theta_j$ are the thresholds for the $J-1 = 4$ levels, $u(\text{judge}_i)$ are judge-specific random effects, and $\beta_1$ and $\beta_2$ are fixed effect coefficients for $\text{temp}_i$ and $\text{contact}_i$.
([see part 1 for more details](https://tdunn.ca/posts/2020-03-15-ordinal-regression-in-r-part-1/#theory)).

```{r}
f_rating_null <- rating ~ 1 + (1|judge)
f_rating_contact_temp <- rating ~ 1 + contact + temp + (1|judge)
```

### Null model

We will start with the "null" model, with just thresholds and random effects.
The default priors for this model are:

```{r}
get_prior(f_rating_null, data = wine, family = cumulative("logit"))
```

$$
\begin{align}
\text{logit}(p(y_i \leq j)) &= \theta_j - u( \text{judge}_i) \\
\theta_j &\sim \text{Student-}t(3, 0, 2.5) \\
u(\text{judge}_i) &\sim \text{Normal}(0, \sigma_u) \\
\sigma_u &\sim \text{Student-} t(3, 0, 2.5)
\end{align}
$$

Fit the model:

```{r}
brm_rating_null <-
  brm(
    f_rating_null,
    data = wine,
    family = cumulative("logit"),
    sample_prior = TRUE,
    file = "brm-rating-null"
  )
```

Visualize the two priors (on thresholds/Intercepts, and on SD of judge effects) with `brms:prior_samples`:

```{r fig.height=3, fig.width=5}
prior_draws(brm_rating_null) %>%
  pivot_longer(cols = everything(), names_to = "term", values_to = "value") %>%
  mutate(samples = "model prior samples") %>%
  ggplot(aes(x = value, y = samples, fill = samples)) +
  geom_violin() +
  # Also visualize some random samples from the t-distribution for comparison
  geom_violin(
    data = tibble(
      value = rstudent_t(n = 4000, df = 3, mu = 0, sigma = 2.5),
      samples = "student_t(3, 0, 2.5)", term = "Intercept"
    )
  ) +
  geom_violin(
    data = tibble(
      value = abs(rstudent_t(n = 4000, df = 3, mu = 0, sigma = 2.5)),
      samples = "abs(student_t(3, 0, 2.5))", term = "sd_judge"
    )
  ) +
  labs(y = NULL) +
  facet_wrap(~term, scales = "free_x") +
  coord_cartesian(xlim = c(-15, 15)) +
  theme(legend.position = "none") +
  add_facet_borders()
```

Note that, since a standard deviation can't be negative, `brms` automatically takes the absolute value of the default `student_t(3, 0, 2.5)` prior.

These priors aren't terrible.
For instance, Intercepts (the thresholds $\theta_j$) between -10 and +10 correspond to the following probabilities:

```{r fig.height=2, fig.width=4}
tibble(theta = seq(-10, 10)) %>%
  mutate(p = inv_logit_scaled(theta)) %>%
  ggplot(aes(x = theta, y = p)) +
  geom_line(size = 1)
```

Any prior values outside of this range would essentially zero cumulative probability for a level $\leq j$.

Now that we've thought about our (default) prior assumptions, investigate the chains:

```{r fig.height=6}
plot(brm_rating_null)
```

The Intercept trace plots look good to me.
There are some spikes in `sd_judge__Intercept`, but not enough to be concerning.

Print the model estimates:

```{r}
brm_rating_null
```

The `Rhat` values are also a good sign of model convergence.
Another thing we can do is a posterior predictive check:

Compare the null Bayesian model estimates to the frequentist estimates:

```{r}
# Can't figure out how to extract random effect SDs from a clmm model, use clmm2
clmm2_rating_null <-
  clmm2(
    rating ~ 1, random = judge,
    data = wine, link = "logistic", Hess = TRUE
  )
# Unfortunately, clmm2 doesn't have a broom::tidy() function
summary(clmm2_rating_null) %>%
  coef() %>%
  as_tibble() %>%
  mutate(term = str_c("b_Intercept[", 1:4, "]")) %>%
  bind_rows(
    tibble(
      Estimate = as.numeric(clmm2_rating_null$stDev),
      term = "sd_judge__Intercept"
    )
  ) %>%
  janitor::clean_names() %>%
  left_join(
    broom.mixed::tidyMCMC(brm_rating_null, conf.int = TRUE),
    by = "term"
  ) %>%
  relocate(term) %>%
  # relocate(term, .before = everything()) %>%
  # select(term, everything()) %>%
  mutate(
    pr_z = scales::pvalue(pr_z),
    across(is.numeric, ~round(., 2))
  ) %>%
  gt() %>%
  tab_spanner(
    label = "ordinal::clmm",
    columns = c(estimate.x, std_error, z_value, pr_z)
  ) %>%
  tab_spanner(
    label = "brms::brm",
    columns = c(estimate.y, std.error, conf.low, conf.high)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "")
```

Frequentist estimates are pretty close to the Bayesian estimates with naive priors.

#### Choice of priors

So what are reasonable priors for this data and model?
My go-to resource for this kind of thing is [this page from the stan wiki](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations), but under the "Prior for cutpoints in ordered logit or probit regression", they have a couple suggestions like "uniform priors typically should be ok", but also say "Need to flesh out this section with examples", so not a lot of help there.

I'll consider the following priors on the the thresholds:

$$
\begin{align}
\theta_j &\sim \text{Student-}t(3, 0, 5.0) \\
\theta_j &\sim \text{Student-}t(3, 0, 2.5) \\
\theta_j &\sim \text{Student-}t(3, 0, 1.0) \\
\theta_j &\sim \text{Normal}(0, 5.0) \\
\theta_j &\sim \text{Normal}(0, 2.5) \\
\theta_j &\sim \text{Normal}(0, 1.0) \\
\end{align}
$$

and simulate some corresponding cumulative probabilities:

```{r fig.height=4, fig.width=5}
bind_rows(
  tibble(sigma = c(1, 2.5, 5), dist = "Normal") %>%
    mutate(
      prior = str_c("Normal(0, ", sigma, ")"),
      samples = map(sigma, ~rnorm(2000, mean = 0, sd = .x))
    ),
  tibble(sigma = c(1, 2.5, 5), dist = "Student-t") %>%
    mutate(prior = str_c("Student-t(3, 0, ", sigma, ")"),
           samples = map(sigma, ~rstudent_t(2000, df = 3, mu = 0, sigma = .x)))
  ) %>%
  #mutate(prior = fct_reorder2(prior, sigma, dist)) %>%
  unnest(samples) %>%
  mutate(p = inv_logit_scaled(samples)) %>%
  ggplot(aes(x = p)) +
  geom_histogram(fill = wine_red, binwidth = 0.1) +
  facet_wrap(~prior) +
  scale_y_continuous(expand = c(0, 0)) +
  add_facet_borders()
```

The $\text{Normal}(0, 5)$ and $\text{Student-}t(3, 0, 5)$ priors place most of the samples at the extremes (cumulative probabilities of 0 and 100%).
The $\text{Normal}(0, 2.5)$ and $\text{Student-}t(3, 0, 2.5)$ are sensible default priors because they are fairly uniform in the probability space, although they do have slight peaks at the extremes.
The $\text{Normal}(0, 1)$ and $\text{Student-}t(3, 0, 1)$ prior distributions are bell-shaped around $p = 0.5$, and might be appropriate if we have reason to believe that there will be no extreme probabilities.

As for the scale parameter describing the variance in judge-specific random effects, I've seen the [half-Cauchy distribution recommended](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations):

```{r fig.height=3, fig.width=5}
tibble(scale = c(1, 2.5, 5)) %>%
  crossing(x = seq(0, 10, 0.1)) %>%
  mutate(
    prior = str_c("Half-Cauchy(0, ", scale, ")"),
    dens = dcauchy(x, location = 0, scale = scale)
  ) %>%
  ggplot(aes(x, y = dens)) +
  geom_line(aes(color = prior), size = 1) +
  theme(legend.position = c(0.6, 0.6), axis.ticks.y = element_blank(),
        axis.text.y = element_blank())
```

This distribution is fairly conservative, with a long right tail that allows for large values.

I don't think there is anything wrong with the default priors, but just to show how it is done, I'll fit the null model with the following:

$$
\begin{align}
\text{logit}(p(y_i \leq j)) &= \theta_j - u( \text{judge}_i) \\
\theta_j &\sim \text{Normal}(0, 1.5) \\
u(\text{judge}_i) &\sim \text{Normal}(0, \sigma_u) \\
\sigma_u &\sim \text{Half-Cauchy}(0, 2.5)
\end{align}
$$

```{r fig.height=5, fig.width=6}
prior_rating_null <- c(
  prior(normal(0, 1.5), class = Intercept),
  prior(cauchy(0, 2.5), class = sd)
)
brm_rating_null_alt_prior <-
  brm(
    f_rating_null,
    prior = prior_rating_null ,
    data = wine,
    family = cumulative("logit"),
    file = "brm-rating-null-alt-prior"
  )
plot(brm_rating_null_alt_prior)
```

As expected, these priors don't noticeably improve the model convergence (which was already good).
Likewise, the model estimates changed only slightly:

```{r}
tidyMCMC(brm_rating_null) %>%
  mutate(priors = "default priors") %>%
  bind_rows(
    tidyMCMC(brm_rating_null_alt_prior) %>%
      mutate(priors = "alternative priors")
  ) %>%
  filter(!str_starts(term, "r_|prior_|disc")) %>%
  transmute(
    term,
    estimate_se = str_c(round(estimate, 2), " (", round(std.error, 2), ")"),
    priors
  ) %>%
  pivot_wider(names_from = priors, values_from = estimate_se) %>%
  gt()
```

### Fixed effects

We now add the "treatment" effects of `temp` and `contact`:

$$
\begin{align}
\text{logit}(p(y_i \leq j)) &= \theta_j - \beta_1 \text{temp}_i - \beta_2 \text{contact}_i - u( \text{judge}_i) \\
\end{align}
$$

This introduces two new priors we can specify:

```{r}
get_prior(f_rating_contact_temp, data = wine, family = cumulative("logit"))
```

We know from part 1 that `contactyes` ($\beta_1$) and `tempwarm` ($\beta_2$) are associated with higher ratings, but we shouldn't be biasing our priors by using the same data we are modeling.
Instead, use a weakly regularizing normal distributions centered at 0:

$$
\begin{align}
\text{logit}(p(y_i \leq j)) &= \theta_j - \beta_1 \text{temp}_i - \beta_2 \text{contact}_i - u( \text{judge}_i) \\
\beta_1 &\sim \text{Normal}(0, 5) \\
\beta_2 &\sim \text{Normal}(0, 5) \\
\theta_j &\sim \text{Normal}(0, 2) \\
u(\text{judge}_i) &\sim \text{Normal}(0, \sigma_u) \\
\sigma_u &\sim \text{Half-Cauchy}(0, 2.5)
\end{align}
$$

```{r}
prior_rating_contact_temp <-
  c(prior_rating_null,
    prior(normal(0, 5), class = b))
brm_rating_contact_temp <-
  brm(
    f_rating_contact_temp,
    prior = prior_rating_contact_temp,
    data = wine,
    family = cumulative("logit"),
    file = "brm_rating_contact_temp_weak_prior"
  )
# Also fit using the default priors
brm_rating_contact_temp_default_prior <-
  brm(
    f_rating_contact_temp,
    data = wine,
    family = cumulative("logit"),
    sample_prior = TRUE,
    file = "brm_rating_contact_temp_default_prior"
  )
brm_rating_contact_temp
```

Compare these estimates to those from `clmm` and with default priors:

```{r}
clmm2_rating_contact_temp <-
  clmm2(
    rating ~ contact + temp, random = judge,
    data = wine, link = "logistic", Hess = TRUE
  )
tidy(brm_rating_contact_temp) %>%
  mutate(model = "brm weak priors") %>%
  bind_rows(
    tidy(brm_rating_contact_temp_default_prior) %>%
      mutate(model = "brm default priors")
  ) %>%
  filter(!str_detect(term, "r_judge|lp__|prior")) %>%
  transmute(
    term, model,
    estimate_se = str_c(round(estimate, 2), " (", round(std.error, 2), ")")
  ) %>%
  bind_rows(
    summary(clmm2_rating_contact_temp) %>%
      coef() %>%
      as_tibble() %>%
      mutate(term = c(str_c("b_Intercept[", 1:4, "]"),
                      "b_contactyes", "b_tempwarm")) %>%
      bind_rows(
        tibble(
          Estimate = as.numeric(clmm2_rating_contact_temp$stDev),
          term = "sd_judge__Intercept"
        )
      ) %>%
      janitor::clean_names() %>%
      transmute(
        model = "clmm", term,
        estimate_se = ifelse(
          !is.na(std_error),
          str_c(round(estimate, 2), " (", round(std_error, 2), ")"),
          round(estimate, 2)
        )
      )
  ) %>%
  spread(model, estimate_se) %>%
  gt()
```

```{r eval=FALSE, include=FALSE}
# Check chain convergence
plot(brm_rating_contact_temp)
plot(brm_rating_contact_temp_weak_prior)
```


## Aside: adjacent-category models

Here are what @Burkner2019 had to say about the adjacent-category class of ordinal models.

* Predict the decision between two adjacent categories $k$ and $k+1$
* Latent variables $\tilde{Y}_k$ with thresholds $\tau_k$ and cumulative distribution function $F$
* If $\tilde{Y}_k < \tau_k$, we choose category $k$; $k+1$ otherwise
* It is difficult to think of a natural process leading to them; chosen for its mathematical convenience rather than quality of interpretation

Mathematically:

$$
\text{Pr}(Y = k | Y \in \{k, k+1\}) = F(\tau_k)
$$

Suppose the latent variable $\tilde{Y}_2$ is standard normally distributed with distribution function $\Phi$, and $\tau_2$ = 1.
Then the probability of choosing $Y$ = 2 over $Y$ = 3 would be written as:

$$
\text{Pr}(Y = 2 | Y \in \{2, 3\}) = \Phi(\tau_2) = \Phi(1) = 0.84
$$
Try fitting the null wine rating model with the `acat` family in `brms`:

```{r}
brm_rating_contact_temp_default_prior_acat <-
  brm(
    f_rating_contact_temp,
    family = acat(link = "logit"),
    data = wine,
    file = "brm_rating_contact_temp_default_prior_acat"
  )
```
How do these estimates compare to the cumulative model?

```{r}
tidy(brm_rating_contact_temp_default_prior) %>%
  mutate(model = "cumulative") %>%
  bind_rows(
    tidy(brm_rating_contact_temp_default_prior_acat) %>%
      mutate(model = "adjacent-category")
  ) %>%
  filter(!str_detect(term, "r_judge|lp__|prior")) %>%
  transmute(
    term, model,
    estimate_se = str_c(round(estimate, 2), " (", round(std.error, 2), ")")
  ) %>%
  spread(model, estimate_se) %>%
  gt()
```

`brms` has a convenience function `marginal_effects` for quickly plotting estimates.
For example, the effect of `category`:

```{r}
me_rating_contact_temp_default_prior_acat <-
  marginal_effects(brm_rating_contact_temp_default_prior_acat, categorical = T,
                   re_formula = NA, ask = F)
me_rating_contact_temp_default_prior <-
  marginal_effects(brm_rating_contact_temp_default_prior, categorical = T,
                   re_formula = NA, ask = F)
me_rating_contact_temp_default_prior_acat$`contact:cats__` %>%
  mutate(model = "adjacent-category") %>%
  bind_rows(
    me_rating_contact_temp_default_prior$`contact:cats__` %>%
      mutate(model = "cumulative")
  ) %>%
  ggplot(aes(x = contact, y = estimate__, color = effect2__)) +
  geom_point(position = position_dodge(1), size = 3) +
  geom_linerange(aes(ymin = lower__, ymax = upper__),
                 position = position_dodge(1), size = 1) +
  facet_wrap(~model) +
  scale_color_viridis_d() +
  labs(y = "Estimated probabilities", color = "rating")
```

The model classes are essentially the same.

## Category-specific effects

In all of the models specified so far, all fixed effects were presumed to affect all response categories equally.
For example, the effect of `temp` = warm had a mean effect of $\beta_1$ =
`r tidy(brm_rating_contact_temp) %>% filter(term == "b_tempwarm") %>% pull(estimate) %>% round(2)` on the thresholds $\theta_j$, for all $j = 1, 2, 3, 4$.

This may not be an appropriate assumption.
For example, `temp` warm might have little relation to the highest `rating`, but it may strongly predict `rating`s of 3 relative to 1 or 2.

If this is a possibility, then we can model the predictor as having a *category-sepcific* effect by estimating $K-1$ coefficients for it.
The reason we've introduced the adjacent-category model is that it is unproblematic to incorporate these effects (sequential models work as well).
Cumulative models, however, can lead to negative probabilities, and so should be avoided when using category-specific effects.

Fit the adjacent-category model with category-specific effects on `temp`:

$$
\begin{align}
\text{logit}(p(y_i \leq j)) &= \theta_j - \beta_{1j} \text{temp}_i - \beta_2 \text{contact}_i - u( \text{judge}_i) \\
\text{logit}(p(y_i \leq j)) &= \theta_j - \beta_{1j} \text{temp}_i - u( \text{judge}_i) \\
\end{align}
$$

```{r}
f_rating_cs_temp <- rating ~ 1 + cs(temp) + (1|judge)
# Note that category-specific priors aren't available
#get_prior(f_rating_cs_temp, data = wine, family = acat(link = "logit"))
brm_rating_cs_temp_default_prior_acat <-
  brm(
    f_rating_cs_temp,
    family = acat(link = "logit"),
    data = wine,
    file = "brm_rating_cs_temp_default_prior_acat"
  )
```

There were many divergent transitions.
Investigate the trace plots:

```{r}
plot(brm_rating_cs_temp_default_prior_acat)
```

Most of the divergence is coming from estimating the lowest and highest coefficients $\beta_{11}$ and $\beta_{14}$.
We will try some regularizing priors and increasing the `adapt_delta` argument:

```{r}
brm_rating_cs_temp_weak_prior_acat <-
  brm(
    f_rating_cs_temp,
    prior = prior_rating_contact_temp,
    family = acat(link = "logit"),
    data = wine,
    file = "brm_rating_cs_temp_weak_prior_acat",
    control = list(adapt_delta = 0.9)
  )
plot(brm_rating_cs_temp_weak_prior_acat)
```

This makes a huge difference.
Compare to a model without category-specific effects:

```{r}
me_rating_cs_temp_weak_prior_acat <-
  marginal_effects(brm_rating_cs_temp_weak_prior_acat, categorical = T,
                   re_formula = NA, ask = F)
brm_rating_temp_weak_prior_acat <-
  brm(
    rating ~ 1 + temp + (1|judge),
    prior = prior_rating_contact_temp,
    family = acat(link = "logit"),
    data = wine,
    file = "brm_rating_temp_weak_prior_acat"
  )
me_rating_temp_weak_prior_acat <-
  marginal_effects(brm_rating_temp_weak_prior_acat, categorical = T,
                   re_formula = NA, ask = F)
me_rating_temp_weak_prior_acat$`temp:cats__` %>%
  mutate(model = "constant effects") %>%
  bind_rows(
    me_rating_cs_temp_weak_prior_acat$`temp:cats__` %>%
      mutate(model = "category-specific effects")
  ) %>%
  ggplot(aes(x = temp, y = estimate__, color = effect2__)) +
  geom_point(position = position_dodge(1), size = 3) +
  geom_linerange(aes(ymin = lower__, ymax = upper__),
                 position = position_dodge(1), size = 1) +
  facet_wrap(~model) +
  scale_color_viridis_d() +
  labs(y = "Estimated probabilities", color = "rating")
```

Or, put the probabilities for each model side-by-side, along with the empirical probabilities:

```{r}
me_rating_temp_weak_prior_acat$`temp:cats__` %>%
  mutate(model = "constant effects") %>%
  bind_rows(
    me_rating_cs_temp_weak_prior_acat$`temp:cats__` %>%
      mutate(model = "category-specific effects")
  ) %>%
  ggp(aes(x = effect2__, y = estimate__, color = model)) +
  geom_point(position = position_dodge(0.5), size = 3) +
  geom_linerange(aes(ymin = lower__, ymax = upper__),
                 position = position_dodge(0.5), size = 1) +
  geom_point(
    data = wine %>%
      group_by(temp, rating) %>%
      tally() %>%
      group_by(temp) %>%
      mutate(p = n / sum(n), model = "empirical"),
    aes(x = rating, y = p, color = model), size = 3
  ) +
  facet_wrap(~temp, ncol = 1) +
  labs(y = "Estimated probabilities", color = "model")
```

We shouldn't be surprised the the category-specific model is better -- it has 4 more parameters to work with -- but we can compare them using leave-one-out cross-validation to determine how much better:

```{r}
brm_rating_temp_weak_prior_acat <-
  add_criterion(brm_rating_temp_weak_prior_acat, "waic")
brm_rating_cs_temp_weak_prior_acat <-
  add_criterion(brm_rating_cs_temp_weak_prior_acat, "waic")
loo_compare(
  brm_rating_temp_weak_prior_acat,
  brm_rating_cs_temp_weak_prior_acat,
  criterion = "waic"
)
```

## Reproducibility {.appendix}

<details><summary>Session info</summary>

```{r echo=FALSE}
devtools::session_info()$platform
devtools::session_info()$packages %>%
  rmarkdown::paged_table()
```

</details>

<details><summary>Git repository</summary>

```{r echo=FALSE}
git2r::repository()
```

</details>

```{r echo=FALSE}
dunnr::get_distill_source(date = params$date, slug = params$slug)
```

