---
title: "Advent of Code 2021: Days 16-20"
description: |
  My solutions to the #AdventOfCode2021 coding challenges, days 16 through 20.
author:
  - name: Taylor Dunn
date: 2021-12-16
params:
  date: 2021-12-16
  slug: "advent-of-code-2021-days-16-20"
categories:
  - Advent of Code
  - Python
  - R
output:
  distill::distill_article:
    self_contained: false
    toc: true
#bibliography: references.bib
---

```{r setup, include=TRUE, code_folding="R setup"}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(gt)
library(tictoc)
#library(lubridate)
```

<details><summary>Python setup</summary>

```{r py-setup1}
Sys.setenv(
  RETICULATE_PYTHON = here("..", "python", ".venv", "Scripts", "python.exe")
)
library(reticulate)
```

```{python py-setup2}
import numpy as np
import pandas as pd
import timeit
#import matplotlib.pylab as plt
#plt.switch_backend('agg') # needed for knitting in rmarkdown
```

</details>

## Day 16: Packet Decoder

```{r}
day16 <- read_lines(here("_posts", "2021-12-16-advent-of-code-2021-days-16-20",
                         "day16-input.txt"))
day16 %>% str_trunc(80)
```

### Part 1

<details><summary>Very long puzzle statement</summary>

>
As you leave the cave and reach open waters, you receive a transmission from the Elves back on the ship.
>
The transmission was sent using the Buoyancy Interchange Transmission System (BITS), a method of packing numeric expressions into a binary sequence. Your submarine's computer has saved the transmission in hexadecimal (your puzzle input).
>
The first step of decoding the message is to convert the hexadecimal representation into binary. Each character of hexadecimal corresponds to four bits of binary data:
>
The BITS transmission contains a single packet at its outermost layer which itself contains many other packets. The hexadecimal representation of this packet might encode a few extra 0 bits at the end; these are not part of the transmission and should be ignored.
>
Every packet begins with a standard header: the first three bits encode the packet version, and the next three bits encode the packet type ID. These two values are numbers; all numbers encoded in any packet are represented as binary with the most significant bit first. For example, a version encoded as the binary sequence 100 represents the number 4.
>
Packets with type ID 4 represent a literal value. Literal value packets encode a single binary number. To do this, the binary number is padded with leading zeroes until its length is a multiple of four bits, and then it is broken into groups of four bits. Each group is prefixed by a 1 bit except the last group, which is prefixed by a 0 bit. These groups of five bits immediately follow the packet header.
>
Every other type of packet (any packet with a type ID other than 4) represent an operator that performs some calculation on one or more sub-packets contained within. Right now, the specific operations aren't important; focus on parsing the hierarchy of sub-packets.
>
An operator packet contains one or more packets. To indicate which subsequent binary data represents its sub-packets, an operator packet can use one of two modes indicated by the bit immediately after the packet header; this is called the length type ID:
>
* If the length type ID is 0, then the next 15 bits are a number that represents the total length in bits of the sub-packets contained by this packet.
* If the length type ID is 1, then the next 11 bits are a number that represents the number of sub-packets immediately contained by this packet.
>
Finally, after the length type ID bit and the 15-bit or 11-bit field, the sub-packets appear.
>
For now, parse the hierarchy of the packets throughout the transmission and add up all of the version numbers.
>
Decode the structure of your hexadecimal-encoded BITS transmission; what do you get if you add up the version numbers in all packets?

</details>

Start by defining the hexadecimal character mapping and converting the input:

```{r}
hex_map <- c(
  "0" = "0000", "1" = "0001", "2" = "0010", "3" = "0011", "4" = "0100",
  "5" = "0101", "6" = "0110", "7" = "0111", "8" = "1000", "9" = "1001",
  "A" = "1010", "B" = "1011", "C" = "1100", "D" = "1101", "E" = "1110",
  "F" = "1111"
)
input <- str_replace_all(day16, hex_map)
str_trunc(input, 90)
```

Second, define a function for parsing literal values (packets with type ID `4`):

```{r}
decode_literal_value <- function(input, i) {
  literal_value <- ""
  repeat {
    lv <- str_sub(input, i, i + 4)
    literal_value <- paste0(literal_value, str_sub(lv, 2, 5))
    i <- i + 5
    # If the bit starts with 0, end of packet
    if (str_sub(lv, 1, 1) == "0") break
    # Else not the last group, keep reading
  }
  
  list(i, literal_value)
}
```

Now define the recursive function to parse the full strings:

```{r}
decode_packet_part1 <- function(input, i = 1) {
  packet_vers <- c(str_sub(input, i, i + 2) %>% strtoi(base = 2))
  packet_type <- str_sub(input, i + 3, i + 5) %>% strtoi(base = 2)
  i <- i + 6
  
  # Literal values
  if (packet_type == 4) {
    literal_value <- decode_literal_value(input, i)
    i <- literal_value[[1]]
    value <- strtoi(literal_value[[2]], base = 2)
    return(list(value = value, index = i, packet_vers = packet_vers))
  }
  
  # Otherwise an operator
  values <- c()
  length_type <- str_sub(input, i, i)
  i <- i + 1
  
  if (length_type == "0") {
    # Next 15 bits are the number of bits in the sub-packets
    n_bits <- str_sub(input, i, i + 14) %>% strtoi(base = 2)
    i <- i + 15
    
    end_bit <- i + n_bits
    while (i < end_bit) {
      subpacket <- decode_packet_part1(input, i)
      packet_vers <- c(packet_vers, subpacket$packet_vers)
      values <- c(values, subpacket$value)
      i <- subpacket$index
    }
  } else if (length_type == "1") {
    # Next 11 bits are the number of sub-packets
    n_subpackets <- str_sub(input, i, i + 10) %>% strtoi(base = 2)
    i <- i + 11
    
    for (sp in 1:n_subpackets) {
      subpacket <- decode_packet_part1(input, i)
      packet_vers <- c(packet_vers, subpacket$packet_vers)
      values <- c(values, subpacket$value)
      i <- subpacket$index
    }  
  }
  
  return(list(value = values, index = i, packet_vers = packet_vers))
}
```

Apply it to the input and sum up the versions:

```{r}
part1_results <- decode_packet_part1(input)
sum(part1_results$packet_vers)
```

Python:

```{python}
hex_map = {'0': '0000', '1': '0001', '2': '0010', '3': '0011', '4': '0100',
           '5': '0101', '6': '0110', '7': '0111', '8': '1000', '9': '1001',
           'A': '1010', 'B': '1011', 'C': '1100', 'D': '1101', 'E': '1110',
           'F': '1111'}
           
input = r.day16
for key, val in hex_map.items():
  input = input.replace(key, val)
  
def decode_literal_value(input, i = 0):
  literal_value = ""
  while True:
    lv = input[i:i+5]
    literal_value += lv[1:]
    i += 5
    
    if lv[0] == "0": break
    
  return i, int(literal_value, 2)

sum_packet_vers = 0
def decode_packet_part1(input, i = 0):
  global sum_packet_vers
  sum_packet_vers += int(input[i:i+3], 2)
  packet_type = int(input[i+3:i+6], 2)
  i += 6
  
  if packet_type == 4:
    i, value = decode_literal_value(input, i)
    return i, value
  
  values = []
  length_type = input[i]
  i += 1
  
  if length_type == "0":
    n_bits = int(input[i:i+15], 2)
    i += 15
    end_bit = i + n_bits
    while i < end_bit:
      i, value = decode_packet_part1(input, i)
      values.append(value)
  else:
    n_subpackets = int(input[i:i+11], 2)
    i += 11
    for _ in range(n_subpackets):
      i, value = decode_packet_part1(input, i)
      values.append(value)
  
  return i, values

_, _ = decode_packet_part1(input)
sum_packet_vers
```

### Part 2

>
Now that you have the structure of your transmission decoded, you can calculate the value of the expression it represents.
>
Literal values (type ID `4`) represent a single number as described above. The remaining type IDs are more interesting:
>
* Packets with type ID `0` are sum packets - their value is the sum of the values of their sub-packets. If they only have a single sub-packet, their value is the value of the sub-packet.
* Packets with type ID `1` are product packets - their value is the result of multiplying together the values of their sub-packets. If they only have a single sub-packet, their value is the value of the sub-packet.
* Packets with type ID `2` are minimum packets - their value is the minimum of the values of their sub-packets.
* Packets with type ID `3` are maximum packets - their value is the maximum of the values of their sub-packets.
* Packets with type ID `5` are greater than packets - their value is `1` if the value of the first sub-packet is greater than the value of the second sub-packet; otherwise, their value is `0`. These packets always have exactly two sub-packets.
* Packets with type ID `6` are less than packets - their value is `1` if the value of the first sub-packet is less than the value of the second sub-packet; otherwise, their value is `0`. These packets always have exactly two sub-packets.
* Packets with type ID `7` are equal to packets - their value is `1` if the value of the first sub-packet is equal to the value of the second sub-packet; otherwise, their value is `0`. These packets always have exactly two sub-packets.
>
Using these rules, you can now work out the value of the outermost packet in your BITS transmission.

Define a function to compute packet values following the above rules:

```{r}
compute_packet_val <- function(packet_type, literal_values) {
  switch(
    packet_type + 1,
    sum(literal_values),
    prod(literal_values),
    min(literal_values),
    max(literal_values),
    NA, # packet_type = 4
    ifelse(literal_values[1] > literal_values[2], 1, 0),
    ifelse(literal_values[1] < literal_values[2], 1, 0),
    ifelse(literal_values[1] == literal_values[2], 1, 0),
  )
}
```

The solution to this part took about an hour of debugging before I realized that the base R function `strtoi`
[cannot handle numbers greater than 2^31](https://stackoverflow.com/questions/13536832/strtoi-fails-to-convert-string-to-integer-returns-na).
Define a custom `strtoi` that accounts for integer overflow:

```{r}
strtoi_custom <- function(x) {
  y <- as.numeric(strsplit(x, "")[[1]])
  sum(y * 2^rev((seq_along(y) - 1)))
}
```

Now re-define the function from part 1 to return the value of the outermost packet in the transmission:

```{r}
decode_packet_part2 <- function(input, i = 1) {
  packet_vers <- c(str_sub(input, i, i + 2) %>% strtoi(base = 2))
  packet_type <- str_sub(input, i + 3, i + 5) %>% strtoi(base = 2)
  i <- i + 6
  
  if (packet_type == 4) {
    literal_value <- decode_literal_value(input, i)
    i <- literal_value[[1]]
    value <- strtoi_custom(literal_value[[2]])
    return(list(value = value, index = i, packet_vers = packet_vers))
  }
  
  # Otherwise, an operator
  values <- c()
  length_type <- str_sub(input, i, i)
  i <- i + 1
  
  if (length_type == "0") {
    # Next 15 bits are the number of bits in the sub-packets
    n_bits <- str_sub(input, i, i + 14) %>% strtoi(base = 2)
    i <- i + 15
    
    end_bit <- i + n_bits
    while (i < end_bit) {
      subpacket <- decode_packet_part2(input, i)
      packet_vers <- c(packet_vers, subpacket$packet_vers)
      values <- c(values, subpacket$value)
      i <- subpacket$index
    }
  } else if (length_type == "1") {
    # Next 11 bits are the number of sub-packets
    n_subpackets <- str_sub(input, i, i + 10) %>% strtoi(base = 2)
    i <- i + 11
    
    for (sp in 1:n_subpackets) {
      subpacket <- decode_packet_part2(input, i)
      packet_vers <- c(packet_vers, subpacket$packet_vers)
      values <- c(values, subpacket$value)
      i <- subpacket$index
    }  
  }
  
  value <- compute_packet_val(packet_type, values)
  
  return(list(value = value, index = i, packet_vers = packet_vers))
}

part2_results <- decode_packet_part2(input)
format(part2_results$value, scientific = FALSE)
```

Python:

```{python}
def compute_packet_val(packet_type, values):
  if packet_type == 0:
    return sum(values)
  elif packet_type == 1:
    prod = 1
    for v in values:
      prod *= v
    return prod
  elif packet_type == 2:
    return min(values)
  elif packet_type == 3:
    return max(values)
  elif packet_type == 5:
    return int(values[0] > values[1]) 
  elif packet_type == 6:
    return int(values[0] < values[1]) 
  elif packet_type == 7:
    return int(values[0] == values[1]) 

def decode_packet_part2(input, i = 0):
  packet_vers = int(input[i:i+3], 2)
  packet_type = int(input[i+3:i+6], 2)
  i += 6
  
  if packet_type == 4:
    i, value = decode_literal_value(input, i)
    return i, value
  
  values = []
  length_type = input[i]
  i += 1
  
  if length_type == "0":
    n_bits = int(input[i:i+15], 2)
    i += 15
    end_bit = i + n_bits
    while i < end_bit:
      i, value = decode_packet_part2(input, i)
      values.append(value)
  else:
    n_subpackets = int(input[i:i+11], 2)
    i += 11
    for _ in range(n_subpackets):
      i, value = decode_packet_part2(input, i)
      values.append(value)
  
  value = compute_packet_val(packet_type, values) 
  
  return i, value

_, value = decode_packet_part2(input)
value
```

## Day 17: Trick Shot

```{r}
day17 <- read_lines(here("_posts", "2021-12-16-advent-of-code-2021-days-16-20",
                         "day17-input.txt"))
day17
```

### Part 1

>
You finally decode the Elves' message. HI, the message says. You continue searching for the sleigh keys.
>
Ahead of you is what appears to be a large ocean trench. Could the keys have fallen into it? You'd better send a probe to investigate.
>
The probe launcher on your submarine can fire the probe with any integer velocity in the x (forward) and y (upward, or downward if negative) directions. For example, an initial x,y velocity like 0,10 would fire the probe straight up, while an initial velocity like 10,-1 would fire the probe forward at a slight downward angle.
>
The probe's x,y position starts at 0,0. Then, it will follow some trajectory by moving in steps. On each step, these changes occur in the following order:
>
* The probe's x position increases by its x velocity.
* The probe's y position increases by its y velocity.
* Due to drag, the probe's x velocity changes by 1 toward the value 0; that is, it decreases by 1 if it is greater than 0, increases by 1 if it is less than 0, or does not change if it is already 0.
* Due to gravity, the probe's y velocity decreases by 1.
>
For the probe to successfully make it into the trench, the probe must be on some trajectory that causes it to be within a target area after any step. The submarine computer has already calculated this target area (your puzzle input).
>
Find the initial velocity that causes the probe to reach the highest y position and still eventually be within the target area after any step. What is the highest y position it reaches on this trajectory?

Use `tidyr::extract` and regex to get the bounds of the target area:

```{r}
target_area <- tibble(x = day17) %>%
  extract(x,
          into = c("xmin", "xmax", "ymin", "ymax"),
          regex = "x=(.*)\\.\\.(.*), y=(.*)\\.\\.(.*)",
          convert = TRUE)
target_area
```

Since the $x$ and $y$ velocity are independent, we can work out the correct $y$ velocity.
Due to the way gravity works in this problem, any velocity $v_y > 0$ starting at $y = 0$ will be $-v_y$ when it returns to $y = 0$ on the way down.
For example, for initial $v_y = [1, 5]$ (`vy_start`), these are the steps and velocities when the probe reaches $y = 0$ again:

```{r}
crossing(vy_start = 1:5, step = 1:20) %>%
  # Each step decreases y velocity by 1
  mutate(vy = vy_start - step + 1) %>%
  group_by(vy_start) %>%
  # The y position is the cumulative sum of y velocities up this point
  mutate(y = cumsum(vy)) %>%
  ungroup() %>%
  filter(y == 0)
```

Then it is just a matter of maximizing `vy_start` such that it lands at the lowest possible point in the target area, i.e. `ymin`.
The highest possible `vy_start` is the one that will reach `ymin` in a single step from `y = 0`, i.e. `vy_max = abs(ymin) - 1`.

```{r}
(vy_max <- abs(target_area$ymin) - 1)
```

To calculate the $y$ trajectory for this initial velocity, use `cumsum`:

```{r}
trajectory <- crossing(vy_max, step = 1:1000) %>%
  mutate(vy = vy_max - step + 1, y = cumsum(vy))
trajectory
```

Here is the step where the probe reaches the edge of the target area:

```{r}
(target_step <- trajectory %>% filter(y == target_area$ymin))
```

And here is the maximum height reached by the probe (the answer to part 1):

```{r}
max(trajectory$y)
```

And just to be sure, before answering, I need to verify that there is an initial $x$ velocity, $v_x$, that is *slow* enough (while still being an integer) that the probe will be between `xmin` and `xmax` (`r target_area$xmin` and `r target_area$xmax`), when it reaches the target area.

Due to drag, the $x$ position will be a cumulative sum of decreasing integers.
As I learned on [day 7](../2021-12-06-advent-of-code-2021-days-6-10/index.html#part-2-1),
this is a triangular number with the value:

$$
\sum_{i=1}^n k = \frac{n (n + 1)}{2}.
$$

This means that the maximum $x$ position reached for an initial $x$ velocity $v_x$ is:

$$
x_{\text{max}} = \frac{v_x (v_x + 1)}{2}
$$


We need to find a $v_x$ that reaches the target $x$ between [156, 202] on or before step `r target_step$step` (which is when we reach the target $y$).

```{r}
tibble(vx_start = 1:30) %>%
  mutate(xmax = vx_start * (vx_start + 1) / 2) %>%
  filter(xmax >= target_area$xmin, xmax <= target_area$xmax)
```

So an initial $v_x$ (`vx_start`) of 18 or 19 will reach the target area in fewer than `r target_step$step` steps, which can be shown with the `cumsum` method as well:

```{r}
tibble(vx_start = c(18, 19)) %>%
  crossing(step = 1:target_step$step) %>%
  mutate(vx = vx_start - step + 1,
         # x velocity will never go lower than 0
         vx = ifelse(vx < 0, 0, vx)) %>%
  group_by(vx_start) %>%
  mutate(x = cumsum(vx)) %>%
  ungroup() %>%
  filter(x >= target_area$xmin, x <= target_area$xmax) %>%
  group_by(vx_start, x, vx) %>%
  filter(step == min(step))
```


### Part 2

>
Maybe a fancy trick shot isn't the best idea; after all, you only have one probe, so you had better not miss.
>
To get the best idea of what your options are for launching the probe, you need to find every initial velocity that causes the probe to eventually be within the target area after any step.
>
How many distinct initial velocity values cause the probe to be within the target area after any step?

Find candidate $y$ velocities, which reach the target area at some `step` in their trajectories:

```{r}
vy_candidates <-
  crossing(vy_start = seq(target_area$ymin, vy_max), step = 1:1000) %>%
  mutate(vy = vy_start - step + 1) %>%
  group_by(vy_start) %>%
  mutate(y = cumsum(vy)) %>%
  ungroup() %>%
  filter(between(y, target_area$ymin, target_area$ymax)) %>%
  select(-vy)
vy_candidates
```

There are `r n_distinct(vy_candidates$vy_start)` unique starting $v_y$ values which will reach the $y$ target.
Do the same for the $x$ target:

```{r}
vx_candidates <- 
  crossing(vx_start = 1:target_area$xmax, step = 1:1000) %>%
  mutate(vx = vx_start - step + 1, vx = ifelse(vx < 0, 0, vx)) %>%
  group_by(vx_start) %>%
  mutate(x = cumsum(vx)) %>%
  ungroup() %>%
  filter(between(x, target_area$xmin, target_area$xmax)) %>%
  select(-vx)
vx_candidates
```

There are `r n_distinct(vx_candidates$vx_start)` unique starting $v_x$ values which will reach the $x$ target.

To determine which combination of $v_x$ and $v_y$ are valid, I need to find the combinations that line up in `step`s taken to reach the target:

```{r}
vxy_candidates <- vx_candidates %>%
  inner_join(vy_candidates, by = "step") %>%
  filter(!is.na(vx_start), !is.na(vy_start))
vxy_candidates %>%
  group_by(vx_start, vy_start) %>%
  summarise(steps = str_c(step, collapse = ", "), .groups = "drop")
```

There are 3202 unique combinations of `vx_start` and `vy_start` which reach the target area during one or more steps.

## Day 18: Snailfish

```{r}
day18 <- read_lines(here("_posts", "2021-12-16-advent-of-code-2021-days-16-20",
                         "day18-input.txt"))
head(day18)
```

### Part 1

<details><summary>Long problem statement</summary>

>You descend into the ocean trench and encounter some snailfish. They say they saw the sleigh keys! They'll even tell you which direction the keys went if you help one of the smaller snailfish with his math homework.
>
Snailfish numbers aren't like regular numbers. Instead, every snailfish number is a pair - an ordered list of two elements. Each element of the pair can be either a regular number or another pair.
>
Pairs are written as `[x,y]`, where x and y are the elements within the pair. Here are some example snailfish numbers, one snailfish number per line:
>
`[1,2]`
`[[1,2],3]`
`[9,[8,7]]`
`[[1,9],[8,5]]`
>
This snailfish homework is about addition. To add two snailfish numbers, form a pair from the left and right parameters of the addition operator. For example, `[1,2]` + `[[3,4],5]` becomes `[[1,2],[[3,4],5]]`.
>
There's only one problem: snailfish numbers must always be reduced, and the process of adding two snailfish numbers can result in snailfish numbers that need to be reduced.
>
To reduce a snailfish number, you must repeatedly do the first action in this list that applies to the snailfish number:
>
* If any pair is nested inside four pairs, the leftmost such pair explodes.
* If any regular number is 10 or greater, the leftmost such regular number splits.
>
Once no action in the above list applies, the snailfish number is reduced.
>
During reduction, at most one action applies, after which the process returns to the top of the list of actions. For example, if split produces a pair that meets the explode criteria, that pair explodes before other splits occur.
>
To explode a pair, the pair's left value is added to the first regular number to the left of the exploding pair (if any), and the pair's right value is added to the first regular number to the right of the exploding pair (if any). Exploding pairs will always consist of two regular numbers. Then, the entire exploding pair is replaced with the regular number `0`.
>
To split a regular number, replace it with a pair; the left element of the pair should be the regular number divided by two and rounded down, while the right element of the pair should be the regular number divided by two and rounded up. For example, `10` becomes `[5,5]`, `11` becomes `[5,6]`, `12` becomes `[6,6]`, and so on.
>
The homework assignment involves adding up a list of snailfish numbers (your puzzle input). The snailfish numbers are each listed on a separate line. Add the first snailfish number and the second, then add that result and the third, then add that result and the fourth, and so on until all numbers in the list have been used once.
>
To check whether it's the right answer, the snailfish teacher only checks the magnitude of the final sum. The magnitude of a pair is 3 times the magnitude of its left element plus 2 times the magnitude of its right element. The magnitude of a regular number is just that number.
>
Add up all of the snailfish numbers from the homework assignment in the order they appear. What is the magnitude of the final sum?

</details>

For reading the data in, I'll parse it like JSON with the `jsonlite` package:

```{r}
library(jsonlite)
input <- map(day18, fromJSON, simplifyVector = FALSE)
```

Through lots of trial and error working through the examples, I wrote the following helper functions:

<details><summary>Helper functions</summary>

This gets the nested location and value of the first-occurring exploding pair:

```{r}
get_explode_pair <- function(sn) {
  ei <- purrr::detect_index(sn, ~vec_depth(.x) >= 5)
   
  if (ei > 0) {
    ep <- pluck(sn, ei)
    # Get the indicies that specify the exploding pair
    vd <- vec_depth(sn[[ei]])
    for (d in seq(vd - 1, 2)) {
      ei <- c(ei, detect_index(ep, ~vec_depth(.x) == d))
      ep <- pluck(ep, tail(ei, 1))
    }
    return(list(ei, unlist(ep)))
  } else {
    return(list())
  }
}
ex1 <- fromJSON("[[[[[9,8],1],2],3],4]", simplifyVector = FALSE)
(res <- get_explode_pair(ex1))
```

This finds the left- and right-most values adjacent to an exploding pair:

```{r}
find_adjacent_idx <- function(sn, ei, side = c("L", "R")) {
  if (side == "L") {
    # Remove levels until we find a node to the left
    while (tail(ei, 1) == 1) {
      ei <- head(ei, -1)
      if (length(ei) == 0) return(ei) # no nodes to the left
    }
    ei[length(ei)] <- ei[length(ei)] - 1
    
    # Add levels until we find the right-most regular number
    #while (length(pluck(sn, !!!ei)) > 1) {
    while (is.list(pluck(sn, !!!ei))) {
      ei <- c(ei, length(pluck(sn, !!!ei)))
    }
    
  } else if (side == "R") {
    # Remove levels until we find a node to the right
    ei[length(ei)] <- ei[length(ei)] + 1
    while (is.null(pluck(sn, !!!ei))) {
      ei <- head(ei, -1)
      if (length(ei) == 0) return(ei) # no nodes to the right
      ei[length(ei)] <- ei[length(ei)] + 1
    }
    
    # Add levels until we find the left-most regular number
    while (is.list(pluck(sn, !!!ei))) {
      ei <- c(ei, 1)
    }
  }
  return(ei)
}
(right_idx <- find_adjacent_idx(ex1, res[[1]], side = "R"))
```

This explodes the pair at the given location:

```{r}
explode_pair <- function(sn, res) {
  ei <- res[[1]]
  ep <- res[[2]]
  
  left_idx <- find_adjacent_idx(sn, ei, side = "L")
  if (length(left_idx) > 0) {
    pluck(sn, !!!left_idx) <- pluck(sn, !!!left_idx) + ep[1]
  }
  right_idx <- find_adjacent_idx(sn, ei, side = "R")
  if (length(right_idx) > 0) {
    pluck(sn, !!!right_idx) <- pluck(sn, !!!right_idx) + ep[2]
  }
  
  pluck(sn, !!!ei) <- 0
  return(sn)
}
explode_pair(ex1, res)
```

This finds the index and value of a split:

```{r}
get_split <- function(sn) {
  # Find the first value >= 10 (if it exists)
  sv <- unlist(sn)
  sv <- sv[sv >= 10][1]
  if (is.na(sv)) {
    return(list())
  } else {
    si <- purrr::detect_index(sn, ~any(unlist(.x) == sv))
    sn <- pluck(sn, si)
    #while (length(sn) > 1) {
    while (is.list(sn)) {
      si <- c(si, detect_index(sn, ~any(unlist(.x) == sv)))
      sn <- pluck(sn, tail(si, 1))
    }
  }
  return(list(si, sv))
}
ex2 <- fromJSON("[[[[0,7],4],[15,[0,13]]],[1,1]]", simplifyVector = FALSE)
(res <- get_split(ex2))
```

This splits the value at the given location:

```{r}
split_val <- function(sn, res) {
  sv <- list(floor(res[[2]] / 2), ceiling(res[[2]] / 2))
  si <- res[[1]]
  pluck(sn, !!!si) <- sv
  return(sn)
}
split_val(ex2, res)
```

</details>

Then the main loop to reduce a snailfish number (or two numbers added together) is:

```{r}
reduce_snailfish_num <- function(sn1, sn2 = NULL) {
  if (is.null(sn2)) {
    sn <- sn1
  } else {
    sn <- list(sn1, sn2)
  }
  
  repeat {
    res_explode <- get_explode_pair(sn)
    if (!is_empty(res_explode)) {
      sn <- explode_pair(sn, res_explode)
    } else {
      res_split <- get_split(sn)
      if (!is_empty(res_split)) {
        sn <- split_val(sn, res_split)
      } else {
        break
      }
    }
  }
  sn
}
```

Applying this loop recursively to the input of `r length(input)` snailfish numbers can be done with `purrr::reduce`:

```{r}
reduced_input <- reduce(input, reduce_snailfish_num)
```

Then the function to calculate the magnitudes:

```{r}
calc_magnitude <- function(reduced_input) {
  repeat {
    if (!is.list(reduced_input)) return(reduced_input)
    
    vd <- vec_depth(reduced_input)
    reduced_input <- reduced_input %>%
      map_depth(
        .depth = vd - 2,
        function(x) {
          if (length(x) == 2) 3 * x[[1]] + 2 * x[[2]]
          else x
        }
      ) %>%
      map_depth(.depth = vd - 2, unlist)
  }
}
```

Compute the magnitude of the input:

```{r}
calc_magnitude(reduced_input)
```

### Part 2

>
You notice a second question on the back of the homework assignment:
>
What is the largest magnitude you can get from adding only two of the snailfish numbers?

The input contains 100 snailfish numbers, so we have to consider 99 * 99 = 9801 unique additions (because addition is commutative in this scenario).
I predict long computation time, so break out the big (i.e. parallel) guns:

```{r}
library(furrr)
future::plan(multisession, workers = 6)
```


```{r cache=TRUE}
tic()
part2_results <- crossing(i = 1:length(input), j = 1:length(input)) %>%
  filter(i != j) %>%
  mutate(
    reduced_input = future_map2(
      i, j,
      ~reduce_snailfish_num(c(input[.x], input[.y]))
    ),
    magnitude = future_map_dbl(reduced_input, calc_magnitude)
  )
toc()
```

And the maximum magnitude:

```{r}
max(part2_results$magnitude)
```

## Day 19: Beacon Scanner

```{r}
day19 <- read_lines(here("_posts", "2021-12-16-advent-of-code-2021-days-16-20",
                         "day19-input.txt"))
head(day19, 30)
```

```{r}
input <- read_lines(
"--- scanner 0 ---
404,-588,-901
528,-643,409
-838,591,734
390,-675,-793
-537,-823,-458
-485,-357,347
-345,-311,381
-661,-816,-575
-876,649,763
-618,-824,-621
553,345,-567
474,580,667
-447,-329,318
-584,868,-557
544,-627,-890
564,392,-477
455,729,728
-892,524,684
-689,845,-530
423,-701,434
7,-33,-71
630,319,-379
443,580,662
-789,900,-551
459,-707,401

--- scanner 1 ---
686,422,578
605,423,415
515,917,-361
-336,658,858
95,138,22
-476,619,847
-340,-569,-846
567,-361,727
-460,603,-452
669,-402,600
729,430,532
-500,-761,534
-322,571,750
-466,-666,-811
-429,-592,574
-355,545,-477
703,-491,-529
-328,-685,520
413,935,-424
-391,539,-444
586,-435,557
-364,-763,-893
807,-499,-711
755,-354,-619
553,889,-390
"
)
```

### Part 1

<details><summary>Long problem statement</summary>
text
</details>

```{r}
scanners <-
  split(input[input != ""], cumsum(input == "")[input != ""]) %>%
  map_dfr(
    function(x) {
      enframe(x[-1], name = "i") %>%
        #mutate(scanner = str_remove_all(x[1], "---") %>% str_trim())
        mutate(scanner = str_extract(x[1], "\\d+") %>% as.integer())
    }
  ) %>%
  separate(value, into = c("x", "y", "z"), sep = ",")
scanners
```


## Reproducibility {.appendix}

<details><summary>Session info</summary>

```{r echo=FALSE}
devtools::session_info()$platform
devtools::session_info()$packages %>%
  rmarkdown::paged_table()
```

</details>

<details><summary>Git repository</summary>

```{r echo=FALSE}
git2r::repository()
```

</details>

```{r echo=FALSE}
dunnr::get_distill_source(date = params$date, slug = params$slug)
```

