---
title: "Advent of Code 2021: Days 11-15"
description: |
  My solutions to the #AdventOfCode2021 coding challenges, days 1 through 5.
author:
  - name: Taylor Dunn
date: 2021-12-11
params:
  date: 2021-12-11
  slug: "advent-of-code-2021-days-11-15"
categories:
  - Advent of Code
  - Python
  - R
output:
  distill::distill_article:
    self_contained: false
    toc: true
#bibliography: references.bib
---

```{r setup, include=TRUE, code_folding="R setup"}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(gt)
library(tictoc)
```

<details><summary>Python setup</summary>

```{r py-setup1}
Sys.setenv(
  RETICULATE_PYTHON = here("..", "python", ".venv", "Scripts", "python.exe")
)
library(reticulate)
```

```{python py-setup2}
import numpy as np
import pandas as pd
```

</details>

## Day 11

```{r}
day11 <- read_lines(here("_posts", "2021-12-11-advent-of-code-2021-days-11-15",
                         "day11-input.txt"))
day11
```

### Part 1

>You enter a large cavern full of rare bioluminescent dumbo octopuses! They seem to not like the Christmas lights on your submarine, so you turn them off for now.
>
There are 100 octopuses arranged neatly in a 10 by 10 grid. Each octopus slowly gains energy over time and flashes brightly for a moment when its energy is full. Although your lights are off, maybe you could navigate through the cave without disturbing the octopuses if you could predict when the flashes of light will happen.
>
Each octopus has an energy level - your submarine can remotely measure the energy level of each octopus (your puzzle input).
>
The energy level of each octopus is a value between 0 and 9. You can model the energy levels and flashes of light in steps. During a single step, the following occurs:
>
* First, the energy level of each octopus increases by 1.
* Then, any octopus with an energy level greater than 9 flashes. This increases the energy level of all adjacent octopuses by 1, including octopuses that are diagonally adjacent. If this causes an octopus to have an energy level greater than 9, it also flashes. This process continues as long as new octopuses keep having their energy level increased beyond 9. (An octopus can only flash at most once per step.)
* Finally, any octopus that flashed during this step has its energy level set to 0, as it used all of its energy to flash.
>
Adjacent flashes can cause an octopus to flash on a step even if it begins that step with very little energy.
Given the starting energy levels of the dumbo octopuses in your cavern, simulate 100 steps. How many total flashes are there after 100 steps?

Put the input into a 10x10 matrix:

```{r}
day11_mat <- day11 %>%
  strsplit("") %>%
  unlist() %>%
  as.integer() %>%
  matrix(nrow = 10, byrow = TRUE)
```

And define a function that returns a point and its neighbors:

```{r}
# The 9 neighbors of a point (including diagonals and the point itself)
d_coords <- cbind(c(0, 0, 0, -1, 1, -1, -1, 1, 1),
                  c(0, -1, 1, 0, 0, -1, 1, -1, 1))

get_neighbors <- function(row_col, max_row = 10, max_col = 10) {
  neighbors <- apply(d_coords, 1, function(d) d + row_col) %>% t()
  in_bounds <- (neighbors[,1] <= max_row) & (neighbors[,1] > 0) &
    (neighbors[,2] <= max_col) & (neighbors[,2] > 0)
  neighbors[in_bounds, , drop = FALSE]
}
```

Now loop over the 100 steps and count the flashes:

```{r}
day11_part1 <- day11_mat
n_flashes <- 0

for (step in 1:100) {
  day11_part1 <- day11_part1 + 1
  # This matrix will keep track of which points have flashed in each step
  flashing <- day11_part1 == 10
  
  n_flashes <- n_flashes + sum(flashing)
  neighbors <- apply(which(flashing, arr.ind = TRUE), 1,
                     get_neighbors, simplify = FALSE)
  
  repeat {
    # Keep track of newly flashing points for each iteration of neighbors
    new_flashing <- matrix(FALSE, nrow = 10, ncol = 10)
     
    for (neighbor in neighbors) {
      day11_part1[neighbor] <- day11_part1[neighbor] + 1
      new_flashing <- (!flashing & (day11_part1 == 10)) | new_flashing
    }
    
    if (sum(new_flashing) == 0) break
    else {
      n_flashes <- n_flashes + sum(new_flashing)
      neighbors <- apply(which(new_flashing, arr.ind = TRUE), 1,
                         get_neighbors, simplify = FALSE)
      flashing <- flashing | new_flashing
    }
  }
  day11_part1[day11_part1 >= 10] <- 0
}

n_flashes
```

This solution took me way way too long to get right because of the `simplify` argument in `apply` being `TRUE` by default.
So when I was collecting a list of `neighbors` around a single point:

```{r}
apply(which(matrix(c(T, F, F, F), nrow = 2, ncol = 2), arr.ind = TRUE), 1,
      get_neighbors)
```

I wouldn't get a list as expected, unless setting `simplify = FALSE`:

```{r}
apply(which(matrix(c(T, F, F, F), nrow = 2, ncol = 2), arr.ind = TRUE), 1,
      get_neighbors, simplify = FALSE)
```
About an hour lost, and a lesson learned.


### Part 2

I can get this answer by just modifying the loop slightly:

```{r}
day11_part2 <- day11_mat
flashing <- day11_part2 == 10

step <- 0
while (!all(flashing)) {
  step <- step + 1
  day11_part2 <- day11_part2 + 1
  
  flashing <- day11_part2 == 10
  #n_flashes <- n_flashes + sum(flashing)
  neighbors <- apply(which(flashing, arr.ind = TRUE), 1,
                     get_neighbors, simplify = FALSE)
  
  repeat {
    new_flashing <- matrix(FALSE, nrow = 10, ncol = 10)
     
    for (neighbor in neighbors) {
      day11_part2[neighbor] <- day11_part2[neighbor] + 1
      new_flashing <- (!flashing & (day11_part2 == 10)) | new_flashing
    }
    
    if (sum(new_flashing) == 0) break
    else {
      neighbors <- apply(which(new_flashing, arr.ind = TRUE), 1,
                         get_neighbors, simplify = FALSE)
      flashing <- flashing | new_flashing
    }
  }
  day11_part2[day11_part2 >= 10] <- 0
}
step
```

## Day 12

```{r}
day12 <- read_lines(here("_posts", "2021-12-11-advent-of-code-2021-days-11-15",
                         "day12-input.txt"))
day12
```

### Part 1

>
With your submarine's subterranean subsystems subsisting suboptimally, the only way you're getting out of this cave anytime soon is by finding a path yourself. Not just a path - the only way to know if you've found the best path is to find all of them.
>
Fortunately, the sensors are still mostly working, and so you build a rough map of the remaining caves (your puzzle input).
>
Your goal is to find the number of distinct paths that start at start, end at end, and don't visit small caves more than once. There are two types of caves: big caves (written in uppercase, like A) and small caves (written in lowercase, like b). It would be a waste of time to visit any small cave more than once, but big caves are large enough that it might be worth visiting them multiple times. So, all paths you find should visit small caves at most once, and can visit big caves any number of times.
>
How many paths through this cave system are there that visit small caves at most once?

First, put together a list of all the cave connections:

```{r}
cave_connections <- tibble(paths = day12) %>%
  separate(paths, into = c("cave1", "cave2"), "-") 
cave_connections <- cave_connections %>%
  # By adding the reverse connections, we have all bi-directional paths
  bind_rows(cave_connections %>% rename(cave1 = cave2, cave2 = cave1)) %>%
  # Paths can't end with start, or start with end
  filter(cave1 != "end", cave2 != "start") %>%
  arrange(cave1)
cave_connections
```

Then the starting paths also begin with the `start` cave:

```{r}
starting_paths <- cave_connections %>%
  filter(cave1 == "start") %>%
  rowwise() %>% # this rowwise() grouping is very important to the logic
  transmute(start_cave = cave1, path = list(cave2),
            end_cave = cave2)
starting_paths
```

Lastly, a helper function to identify small caves (I'm not aware of a base R function to check for case):

```{r}
is_lower <- function(x) {
  tolower(x) == x
}
```

My strategy is to continuously loop, add new caves by `left_join`ing `cave_connections`, add new caves to the `path` list, and remove paths which go through the same small cave twice.
The loop will `break` when there was no change from the previous iteration:

```{r}
tic()
cave_paths <- starting_paths
repeat {
  prev_paths <- cave_paths
  
  cave_paths <- cave_paths %>%
    left_join(
      cave_connections, by = c("end_cave" = "cave1")
    ) %>%
    # Remove paths which visit a small cave twice
    filter(
      !(is_lower(cave2) & (cave2 %in% path))
    ) %>%
    # Add the new caves to the paths, unless NA (already at the end)
    mutate(
      path = ifelse(!is.na(cave2), list(c(path, cave2)), list(path)),
      end_cave = ifelse(!is.na(cave2), cave2, end_cave)
    ) %>%
    select(-cave2)
  
  if (all_equal(prev_paths, cave_paths) == TRUE) break 
}
toc()
```

`r n_distinct(cave_paths$path)` unique paths were identified:

```{r}
cave_paths %>%
  mutate(path = str_c(c(start_cave, path), collapse = ",")) %>%
  distinct(path)
```

### Part 2

>
After reviewing the available paths, you realize you might have time to visit a single small cave twice. Specifically, big caves can be visited any number of times, a single small cave can be visited at most twice, and the remaining small caves can be visited at most once. However, the caves named start and end can only be visited exactly once each: once you leave the start cave, you may not return to it, and once you reach the end cave, the path must end immediately.
>
Given these new rules, how many paths through this cave system are there?

I will slightly modify the loop from part 1 to check if a small cave has been visited twice:

```{r part2_loop, cache=TRUE}
tic()
cave_paths <- starting_paths %>% mutate(small_cave_twice = FALSE)
repeat {
  prev_paths <- cave_paths
  
  cave_paths <- cave_paths %>%
    left_join(
      cave_connections, by = c("end_cave" = "cave1")
    ) %>%
    # Remove paths which visit a small cave twice more than once
    filter(
      !(small_cave_twice & is_lower(cave2) & (cave2 %in% path))
    ) %>%
    # Add the new caves to the paths, unless NA (already at the end)
    mutate(
      path = ifelse(!is.na(cave2), list(c(path, cave2)), list(path)),
      end_cave = ifelse(!is.na(cave2), cave2, end_cave)
    ) %>%
    select(-cave2) %>%
    # Check to see if a small cave has been visited twice
    mutate(
      small_cave_twice = any(table(path[is_lower(path)]) > 1)  
    )
  if (all_equal(prev_paths, cave_paths) == TRUE) break 
}
toc()
```

`r n_distinct(cave_paths$path)` unique paths were identified:

```{r}
cave_paths %>%
  mutate(path = str_c(c(start_cave, path), collapse = ",")) %>%
  distinct(path)
```

This solution was obviously inefficient, taking about 2 minutes to run.

>[.. every problem has a solution that completes in at most 15 seconds on ten-year-old hardware.](https://adventofcode.com/2021/about)

## Reproducibility {.appendix}

<details><summary>Session info</summary>

```{r echo=FALSE}
devtools::session_info()$platform
devtools::session_info()$packages %>%
  rmarkdown::paged_table()
```

</details>

<details><summary>Git repository</summary>

```{r echo=FALSE}
git2r::repository()
```

</details>

```{r echo=FALSE}
dunnr::get_distill_source(date = params$date, slug = params$slug)
```

