---
title: "Advent of Code 2021: Days 11-15"
description: |
  My solutions to the #AdventOfCode2021 coding challenges, days 1 through 5.
author:
  - name: Taylor Dunn
date: 2021-12-11
params:
  date: 2021-12-11
  slug: "advent-of-code-2021-days-11-15"
categories:
  - Advent of Code
  - Python
  - R
output:
  distill::distill_article:
    self_contained: false
    toc: true
#bibliography: references.bib
---

```{r setup, include=TRUE, code_folding="R setup"}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(gt)
```

<details><summary>Python setup</summary>

```{r py-setup1}
Sys.setenv(
  RETICULATE_PYTHON = here("..", "python", ".venv", "Scripts", "python.exe")
)
library(reticulate)
```

```{python py-setup2}
import numpy as np
import pandas as pd
```

</details>

## Day 11

```{r}
day11 <- read_lines(here("_posts", "2021-12-11-advent-of-code-2021-days-11-15",
                         "day11-input.txt"))
day11
```

### Part 1

>You enter a large cavern full of rare bioluminescent dumbo octopuses! They seem to not like the Christmas lights on your submarine, so you turn them off for now.
>
There are 100 octopuses arranged neatly in a 10 by 10 grid. Each octopus slowly gains energy over time and flashes brightly for a moment when its energy is full. Although your lights are off, maybe you could navigate through the cave without disturbing the octopuses if you could predict when the flashes of light will happen.
>
Each octopus has an energy level - your submarine can remotely measure the energy level of each octopus (your puzzle input).
>
The energy level of each octopus is a value between 0 and 9. You can model the energy levels and flashes of light in steps. During a single step, the following occurs:
>
* First, the energy level of each octopus increases by 1.
* Then, any octopus with an energy level greater than 9 flashes. This increases the energy level of all adjacent octopuses by 1, including octopuses that are diagonally adjacent. If this causes an octopus to have an energy level greater than 9, it also flashes. This process continues as long as new octopuses keep having their energy level increased beyond 9. (An octopus can only flash at most once per step.)
* Finally, any octopus that flashed during this step has its energy level set to 0, as it used all of its energy to flash.
>
Adjacent flashes can cause an octopus to flash on a step even if it begins that step with very little energy.
Given the starting energy levels of the dumbo octopuses in your cavern, simulate 100 steps. How many total flashes are there after 100 steps?

Put the input into a 10x10 matrix:

```{r}
day11_mat <- day11 %>%
  strsplit("") %>%
  unlist() %>%
  as.integer() %>%
  matrix(nrow = 10, byrow = TRUE)
```

And define a function that returns a point and its neighbors:

```{r}
# The 9 neighbors of a point (including diagonals and the point itself)
d_coords <- cbind(c(0, 0, 0, -1, 1, -1, -1, 1, 1),
                  c(0, -1, 1, 0, 0, -1, 1, -1, 1))

get_neighbors <- function(row_col, max_row = 10, max_col = 10) {
  neighbors <- apply(d_coords, 1, function(d) d + row_col) %>% t()
  in_bounds <- (neighbors[,1] <= max_row) & (neighbors[,1] > 0) &
    (neighbors[,2] <= max_col) & (neighbors[,2] > 0)
  neighbors[in_bounds, , drop = FALSE]
}
```

Now loop over the 100 steps and count the flashes:

```{r}
day11_part1 <- day11_mat
n_flashes <- 0

for (step in 1:100) {
  day11_part1 <- day11_part1 + 1
  # This matrix will keep track of which points have flashed in each step
  flashing <- day11_part1 == 10
  
  n_flashes <- n_flashes + sum(flashing)
  neighbors <- apply(which(flashing, arr.ind = TRUE), 1,
                     get_neighbors, simplify = FALSE)
  
  repeat {
    # Keep track of newly flashing points for each iteration of neighbors
    new_flashing <- matrix(FALSE, nrow = 10, ncol = 10)
     
    for (neighbor in neighbors) {
      day11_part1[neighbor] <- day11_part1[neighbor] + 1
      new_flashing <- (!flashing & (day11_part1 == 10)) | new_flashing
    }
    
    if (sum(new_flashing) == 0) break
    else {
      n_flashes <- n_flashes + sum(new_flashing)
      neighbors <- apply(which(new_flashing, arr.ind = TRUE), 1,
                         get_neighbors, simplify = FALSE)
      flashing <- flashing | new_flashing
    }
  }
  day11_part1[day11_part1 >= 10] <- 0
}

n_flashes
```

This solution took me way way too long to get right because of the `simplify` argument in `apply` being `TRUE` by default.
So when I was collecting a list of `neighbors` around a single point:

```{r}
apply(which(matrix(c(T, F, F, F), nrow = 2, ncol = 2), arr.ind = TRUE), 1,
      get_neighbors)
```

I wouldn't get a list as expected, unless setting `simplify = FALSE`:

```{r}
apply(which(matrix(c(T, F, F, F), nrow = 2, ncol = 2), arr.ind = TRUE), 1,
      get_neighbors, simplify = FALSE)
```
About an hour lost, and a lesson learned.

### Part 2

I can get this answer by just modifying the loop slightly:

```{r}
day11_part2 <- day11_mat
flashing <- day11_part2 == 10

step <- 0
while (!all(flashing)) {
  step <- step + 1
  day11_part2 <- day11_part2 + 1
  
  flashing <- day11_part2 == 10
  #n_flashes <- n_flashes + sum(flashing)
  neighbors <- apply(which(flashing, arr.ind = TRUE), 1,
                     get_neighbors, simplify = FALSE)
  
  repeat {
    new_flashing <- matrix(FALSE, nrow = 10, ncol = 10)
     
    for (neighbor in neighbors) {
      day11_part2[neighbor] <- day11_part2[neighbor] + 1
      new_flashing <- (!flashing & (day11_part2 == 10)) | new_flashing
    }
    
    if (sum(new_flashing) == 0) break
    else {
      neighbors <- apply(which(new_flashing, arr.ind = TRUE), 1,
                         get_neighbors, simplify = FALSE)
      flashing <- flashing | new_flashing
    }
  }
  day11_part2[day11_part2 >= 10] <- 0
}
step
```

# Reproducibility {.appendix}

<details><summary>Session info</summary>

```{r echo=FALSE}
devtools::session_info()$platform
devtools::session_info()$packages %>%
  rmarkdown::paged_table()
```

</details>

<details><summary>Git repository</summary>

```{r echo=FALSE}
git2r::repository()
```

</details>

```{r echo=FALSE}
dunnr::get_distill_source(date = params$date, slug = params$slug)
```

